\documentclass{article}
\usepackage{neurips_2024}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{stmaryrd}       % for \llbracket, \rrbracket

% Theorem environments
\newtheorem{definition}{Definition}
\newtheorem{assumption}{Assumption}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{remark}{Remark}

% Operator shortcuts
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\pre}{pre}
\DeclareMathOperator{\eff}{eff}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\img}{img}
\newcommand{\calS}{\mathcal{S}}
\newcommand{\calA}{\mathcal{A}}
\newcommand{\calO}{\mathcal{O}}
\newcommand{\calC}{\mathcal{C}}
\newcommand{\calD}{\mathcal{D}}
\newcommand{\calE}{\mathcal{E}}
\newcommand{\calF}{\mathcal{F}}
\newcommand{\calG}{\mathcal{G}}
\newcommand{\calH}{\mathcal{H}}
\newcommand{\calI}{\mathcal{I}}
\newcommand{\calK}{\mathcal{K}}
\newcommand{\calL}{\mathcal{L}}
\newcommand{\calM}{\mathcal{M}}
\newcommand{\calN}{\mathcal{N}}
\newcommand{\calP}{\mathcal{P}}
\newcommand{\calR}{\mathcal{R}}
\newcommand{\calT}{\mathcal{T}}
\newcommand{\calU}{\mathcal{U}}
\newcommand{\calV}{\mathcal{V}}
\newcommand{\calW}{\mathcal{W}}
\newcommand{\calX}{\mathcal{X}}
\newcommand{\bbR}{\mathbb{R}}
\newcommand{\bbN}{\mathbb{N}}
\newcommand{\bbZ}{\mathbb{Z}}
\newcommand{\ind}{\mathbb{1}}
\newcommand{\Atoms}{\mathsf{Atoms}}
\newcommand{\Obj}{\mathsf{Obj}}
\newcommand{\Pred}{\mathsf{Pred}}
\newcommand{\strips}{\textsc{Strips}}
\newcommand{\pddl}{\textsc{Pddl}}
\newcommand{\vlm}{\textsc{Vlm}}

\title{Formalism}

\author{}

\begin{document}

\maketitle

\begin{abstract}

\end{abstract}


% ============================================================================
% SECTION 1: PRELIMINARIES
% ============================================================================

\section{Preliminaries}
\label{sec:preliminaries}

We establish the logical and planning-theoretic foundations.

\subsection{First-Order Logic Foundations}
\label{sec:fol}

\begin{definition}[First-Order Language]
\label{def:fol_language}
A \textbf{first-order language} $\calL = (\calC, \calP, \calF)$ consists of:
\begin{itemize}[nosep]
    \item A set of \textbf{constant symbols} $\calC = \{c_1, c_2, \ldots\}$ denoting individual objects in the domain;
    \item A set of \textbf{predicate symbols} $\calP = \calP^{(1)} \cup \calP^{(2)} \cup \cdots$ partitioned by arity, where $\calP^{(k)}$ contains $k$-ary predicate symbols;
    \item A set of \textbf{function symbols} $\calF = \calF^{(1)} \cup \calF^{(2)} \cup \cdots$ partitioned by arity.
\end{itemize}
\end{definition}

In our domain, the language is structured around objects detected in the grid environment.  We distinguish three families of predicates that capture the relevant structure:

\begin{definition}[Domain-Specific Predicate Families]
\label{def:pred_families}
Given a grid environment, we define:
\begin{enumerate}[nosep]
    \item \textbf{Unary predicates} $\calP^{(1)}$: properties of individual objects, e.g., $\mathsf{IsPlayer}(o)$, $\mathsf{IsWall}(o)$, $\mathsf{IsGoal}(o)$, $\mathsf{IsMovable}(o)$, $\mathsf{IsRectangular}(o)$.
    \item \textbf{Binary predicates} $\calP^{(2)}$: relations between pairs of objects, e.g., $\mathsf{Adjacent}(o_1, o_2)$, $\mathsf{Above}(o_1, o_2)$, $\mathsf{Blocking}(o_1, o_2)$, $\mathsf{SameColor}(o_1, o_2)$, $\mathsf{Collides}(o_1, o_2)$.
    \item \textbf{Functional terms} $\calF$: attribute functions returning values, e.g., $\mathsf{color}(o) \in \{0, \ldots, K-1\}$, $\mathsf{pos\_x}(o) \in \bbZ$, $\mathsf{pos\_y}(o) \in \bbZ$, $\mathsf{area}(o) \in \bbN$, $\mathsf{type}(o) \in \Sigma^*$.
\end{enumerate}
\end{definition}

\begin{definition}[First-Order Structure / Interpretation]
\label{def:fo_structure}
A \textbf{first-order structure} (or \textbf{interpretation}) over $\calL$ is a tuple $\calI = (D, \cdot^{\calI})$, where $D$ is a non-empty \textbf{domain} (the set of objects in a game state) and $\cdot^{\calI}$ is an interpretation function mapping each constant $c \in \calC$ to an element $c^{\calI} \in D$, each $k$-ary predicate $P \in \calP^{(k)}$ to a relation $P^{\calI} \subseteq D^k$, and each $k$-ary function $f \in \calF^{(k)}$ to a total function $f^{\calI} : D^k \to D'$ for an appropriate range $D'$.
\end{definition}

% \begin{definition}[Herbrand Base and State as Interpretation]
% \label{def:herbrand}
% The \textbf{Herbrand base} $\calH_{\calL}$ of language $\calL$ with constant set $\calC$ is the set of all ground atoms:
% \begin{equation}
% \calH_{\calL} = \{ P(c_1, \ldots, c_k) : P \in \calP^{(k)},\; c_1, \ldots, c_k \in \calC \}.
% \end{equation}
% A \textbf{logical state} is a subset $s \subseteq \calH_{\calL}$ of ground atoms that are true, under the \textbf{closed-world assumption} (CWA): any ground atom not in $s$ is false.
% \end{definition}

\subsection{STRIPS Planning and Action Models}
\label{sec:strips}

Following the classical AI planning literature~\cite{fikes1971strips, ghallab2004automated}, we define the action model representation that underpins our formalism.

\begin{definition}[STRIPS Action Schema]
\label{def:strips}
A \textbf{STRIPS action schema} is a triple $\alpha = (\pre(\alpha), \eff^+(\alpha), \eff^-(\alpha))$, where:
\begin{itemize}[nosep]
    \item $\pre(\alpha) \subseteq \calH_{\calL}$ is the set of \textbf{preconditions}: ground atoms that must hold for $\alpha$ to be applicable;
    \item $\eff^+(\alpha) \subseteq \calH_{\calL}$ is the \textbf{add list}: atoms that become true after executing $\alpha$;
    \item $\eff^-(\alpha) \subseteq \calH_{\calL}$ is the \textbf{delete list}: atoms that become false after executing $\alpha$.
\end{itemize}
Action $\alpha$ is \textbf{applicable} in state $s$ iff $\pre(\alpha) \subseteq s$.  The result of applying $\alpha$ in $s$ is:
\begin{equation}
\label{eq:strips_transition}
\gamma(s, \alpha) = (s \setminus \eff^-(\alpha)) \cup \eff^+(\alpha).
\end{equation}
\end{definition}

\begin{definition}[PDDL Planning Domain]
\label{def:pddl_domain}
A \textbf{PDDL planning domain}~\cite{mcdermott1998pddl, fox2003pddl2} is a tuple $\calD = (\calT, \calP, \calA_{\text{schema}}, \Gamma)$, where:
\begin{itemize}[nosep]
    \item $\calT$ is a type hierarchy over objects;
    \item $\calP$ is a set of typed predicate symbols;
    \item $\calA_{\text{schema}} = \{\alpha_1, \ldots, \alpha_N\}$ is a finite set of \textbf{lifted action schemas}, each parameterized by typed variables;
    \item $\Gamma$ is the set of ground instances obtained by substituting constants for variables.
\end{itemize}
A \textbf{planning problem} is $\Pi = (\calD, I, G)$, where $I \subseteq \calH_{\calL}$ is the initial state and $G \subseteq \calH_{\calL}$ is the goal condition.
\end{definition}

\begin{remark}[Implicit STRIPS in Our Setting]
\label{rem:implicit_strips}
In the ARC-AGI-3 domain, the agent does not receive action schemas a priori.  Instead, the agent must \emph{learn} the preconditions and effects of each action from interaction, this is the \emph{action model learning} problem~\cite{cresswell2011acquiring, arora2018review, juba2021safe, stern2017efficient}.  The VLM-generated natural language definitions $\{d_a\}_{a \in \calA}$ implicitly encode a STRIPS-like representation:
\begin{equation}
\label{eq:implicit_strips}
d_a \;\text{encodes}\; \bigl(\widehat{\pre}(a),\; \widehat{\eff}^+(a),\; \widehat{\eff}^-(a)\bigr),
\end{equation}
where the hat denotes the agent's current estimate.  The executable Python model $\hat{T}_\theta$ then operationalizes these estimates as code.
\end{remark}

\subsection{Model-Based Reinforcement Learning Notation}
\label{sec:mbrl_notation}

We adopt standard notation from model-based RL (MBRL)~\cite{sutton1991dyna, ha2018world, hafner2020dreamer, schrittwieser2020muzero} while noting departures.

\begin{definition}[Deterministic MDP without Reward]
\label{def:mdp}
An \textbf{interactive environment} is a reward-free deterministic MDP:
\begin{equation}
\label{eq:mdp}
\calM = (\calS, \calA, T, s_0),
\end{equation}
where $\calS$ is the state space, $\calA$ is the finite action set, $T : \calS \times \calA \to \calS$ is the deterministic transition function, and $s_0 \in \calS$ is the initial state.  We deliberately exclude a reward function: the agent's objective is \emph{model acquisition} rather than policy optimization.
\end{definition}

\begin{remark}[Departures from Standard MBRL]
\label{rem:mbrl_departures}
In Dreamer~\cite{hafner2020dreamer} and MuZero~\cite{schrittwieser2020muzero}, world models are learned via gradient descent on observation/reward prediction losses ($\calL = \mathbb{E}[\|f_\theta(s,a) - s'\|^2]$).  Our approach instead uses a VLM as a ``compiler'' that synthesizes $\hat{T}$ as a symbolic program from natural language, bypassing gradient-based optimization entirely.  This is closer to the program synthesis approach of DreamCoder~\cite{ellis2021dreamcoder}, where the model is a composable program rather than a neural network.
\end{remark}

\subsection{PAC-MDP Framework}
\label{sec:pac_mdp}

We connect to the PAC-MDP framework~\cite{kakade2003sample, brafman2002rmax, strehl2009reinforcement} for sample-efficient exploration.

\begin{definition}[PAC-MDP]
\label{def:pac_mdp}
An algorithm is \textbf{PAC-MDP} if, with probability at least $1 - \delta$, the total number of timesteps $t$ where the algorithm's policy is not $\epsilon$-optimal is at most polynomial in $|\calS|$, $|\calA|$, $1/\epsilon$, $1/\delta$, and $H$ (the horizon).
\end{definition}

\begin{definition}[$\epsilon$-Correct Transition Model]
\label{def:eps_correct}
A learned model $\hat{T}$ is \textbf{$\epsilon$-correct} on a set $\calD \subseteq \calS \times \calA$ if:
\begin{equation}
\forall (s, a) \in \calD: \quad \hat{T}(s, a) = T(s, a).
\end{equation}
More generally, $\hat{T}$ is $\epsilon$-correct in the distributional sense if $\Pr_{(s,a) \sim \mu}[\hat{T}(s,a) \neq T(s,a)] \leq \epsilon$.
\end{definition}

The R-MAX algorithm~\cite{brafman2002rmax} achieves PAC-MDP guarantees by maintaining a partition of state-action pairs into \emph{known} and \emph{unknown} sets.  We adapt this structure:

\begin{definition}[Known/Unknown Partition]
\label{def:known_unknown}
At time $t$, we partition $\calS \times \calA$ into:
\begin{align}
\calK_t^{\text{known}} &= \{(s, a) : \text{the agent's model is verified correct on } (s, a)\}, \\
\calK_t^{\text{unknown}} &= (\calS \times \calA) \setminus \calK_t^{\text{known}}.
\end{align}
Following R-MAX~\cite{brafman2002rmax}, the agent acts optimistically on unknown state-action pairs, prioritizing exploration of $\calK_t^{\text{unknown}}$.
\end{definition}


% ============================================================================
% SECTION 2: PROBLEM SETTING
% ============================================================================

\section{Problem Setting}
\label{sec:problem}

\subsection{Grid Environment as First-Order Structure}
\label{sec:grid_env}

\begin{definition}[Grid Environment]
\label{def:grid_env}
An \textbf{interactive grid environment} is specified by $\calM = (\calS, \calA, T, s_0)$ (Definition~\ref{def:mdp}), where:
\begin{itemize}[nosep]
    \item $\calS \subseteq \bbZ_K^{H \times W}$ is the set of reachable states.  Each $s \in \calS$ is an integer-valued grid of height $H$ and width $W$ with $K$ possible cell values (colors): $s_{ij} \in \{0, \ldots, K-1\}$ for all $(i,j) \in [H] \times [W]$.
    \item $\calA = \{a_1, \ldots, a_N\}$ is a finite action set.  The agent knows $|\calA|$ but not the semantics of any $a \in \calA$.
    \item $T : \calS \times \calA \to \calS$ is a deterministic, unknown transition function.
    \item $s_0 \in \calS$ is the initial state, observed at the start of interaction.
\end{itemize}
\end{definition}

\subsection{Object Detection: From Pixels to First-Order Domain}
\label{sec:object_detection}

The bridge from raw grid states to first-order logical representations is the \emph{object detection function}, which constructs the domain $D$ of a first-order structure from pixel data.

\begin{definition}[Object]
\label{def:object}
An \textbf{object} is a tuple $o = (c, P, \phi)$, where:
\begin{itemize}[nosep]
    \item $c \in \{0, \ldots, K-1\}$ is the object's \textbf{color};
    \item $P \subseteq [H] \times [W]$ is a connected\footnote{Connected under 8-connectivity on the integer lattice $\bbZ^2$, following Gestalt principles of proximity and similarity~\cite{wertheimer1938laws}.} set of occupied cells;
    \item $\phi = (\mathsf{bbox}, \mathsf{center}, \mathsf{area}, \mathsf{shape}) \in \calP$ is a vector of derived properties.
\end{itemize}
\end{definition}

\begin{definition}[Connected-Component Object Detection]
\label{def:cc_detect}
The \textbf{object detection function} $\Omega : \calS \to 2^{\calO}$ applies connected-component labeling~\cite{he2017connected} with 8-connectivity to each color channel of the grid.  For state $s$:
\begin{equation}
\Omega(s) = \bigsqcup_{c=1}^{K-1} \textsc{CC}_8\bigl(\{(i,j) : s_{ij} = c\}\bigr),
\end{equation}
where $\textsc{CC}_8(\cdot)$ returns the connected components under 8-adjacency, and $\bigsqcup$ denotes disjoint union.  Color $0$ (background) is excluded.  Each component becomes an object $o_k = (c_k, P_k, \phi_k)$.
\end{definition}

This is implemented in our system via \texttt{scipy.ndimage.label} with a $3 \times 3$ structuring element that captures 8-connectivity.  The detector is deterministic, parameter-free, and domain-agnostic, analogous to the perceptual grouping mechanisms studied in Slot Attention~\cite{locatello2020object} and OP3~\cite{veerapaneni2020entity}, but without learned parameters.

\subsection{Logical State Lifting}
\label{sec:lifting}

Given the detected objects, we lift the raw grid state into a first-order logical state.

\begin{definition}[Predicate Grounding Function]
\label{def:predicate_grounding}
Following Athalye et al.~\cite{athalye2024pixels}, a \textbf{predicate grounding function} is a mapping that associates each predicate symbol with a computable classifier over the raw state:
\begin{equation}
\label{eq:grounding}
\mathsf{ground} : P \in \calP^{(k)} \mapsto g_P : D^k \to \{0, 1\},
\end{equation}
where $g_P(o_1, \ldots, o_k) = 1$ iff the relation $P$ holds among objects $o_1, \ldots, o_k$ in the current state.
\end{definition}

\begin{definition}[Logical State Lifting]
\label{def:lifting}
The \textbf{lifting function} $\Lambda : \calS \to 2^{\calH_\calL}$ maps a raw grid state to a logical state:
\begin{equation}
\Lambda(s) = \{ P(o_1, \ldots, o_k) : P \in \calP^{(k)},\; (o_1, \ldots, o_k) \in \Omega(s)^k,\; g_P(o_1, \ldots, o_k) = 1 \}.
\end{equation}
The \textbf{structured state} is the pair $\bar{s} = (s, \Lambda(s))$, preserving both the raw grid and its logical interpretation.
\end{definition}

\begin{definition}[FOL Transition via STRIPS]
\label{def:fol_transition}
Given the lifting function $\Lambda$, the transition function $T$ induces a \textbf{logical transition function} $T_\Lambda$ on logical states:
\begin{equation}
T_\Lambda(\Lambda(s), a) = \Lambda(T(s, a)).
\end{equation}
If the environment's dynamics are \emph{STRIPS-representable}, i.e., there exist schemas $\alpha_a = (\pre(\alpha_a), \eff^+(\alpha_a), \eff^-(\alpha_a))$ for each $a \in \calA$, then $T_\Lambda$ factors as:
\begin{equation}
T_\Lambda(\Lambda(s), a) = (\Lambda(s) \setminus \eff^-(\alpha_a)) \cup \eff^+(\alpha_a), \quad \text{whenever } \pre(\alpha_a) \subseteq \Lambda(s).
\end{equation}

\end{definition}

\begin{remark}[Relation to Object-Centric World Models]
This lifting connects to C-SWM~\cite{kipf2020contrastive}, which learns object-centric state abstractions for planning, and to Relation Networks~\cite{santoro2017simple}, which reason about pairwise object relations.  Our difference is that predicate grounding is performed by the VLM oracle (Section~\ref{sec:vlm_oracle}) rather than learned from data, enabling zero-shot transfer to novel environments.  The lifting $\Lambda$ induces a natural \textbf{state abstraction} in the sense of Li et al.~\cite{li2006towards}: two raw states $s_1, s_2$ with $\Lambda(s_1) = \Lambda(s_2)$ are \emph{action-equivalent} if for all actions $a$, $\Lambda(T(s_1, a)) = \Lambda(T(s_2, a))$.  This connects to the concept of \emph{bisimulation} over relational MDPs~\cite{givan2003equivalence}.
\end{remark}

\subsection{Interaction Protocol}
\label{sec:protocol}

The agent interacts with $\calM$ in discrete timesteps $t = 0, 1, 2, \ldots$:
\begin{enumerate}[nosep]
    \item \textbf{Observe}: receive $s_t \in \calS$.
    \item \textbf{Act}: select $a_t \in \calA$.
    \item \textbf{Transition}: environment transitions to $s_{t+1} = T(s_t, a_t)$.
    \item \textbf{Observe}: receive $s_{t+1}$.
\end{enumerate}
After $t$ interactions, the agent's trajectory is $\tau_t = \{(s_0, a_0, s_1), \ldots, (s_{t-1}, a_{t-1}, s_t)\}$.  The interaction budget $B \in \bbN$ is finite.

\subsection{Model Acquisition Objective}
\label{sec:objective}

\begin{definition}[Transition Prediction Error]
\label{def:pred_error}
Given $\hat{T} : \calS \times \calA \to \calS$ and a sample $(s, a, s')$ where $s' = T(s, a)$:
\begin{equation}
\epsilon(s, a; \hat{T}) = \frac{1}{HW} \sum_{i=1}^{H} \sum_{j=1}^{W} \ind[\hat{T}(s, a)_{ij} \neq s'_{ij}].
\end{equation}
\end{definition}

\begin{definition}[Model Accuracy]
\label{def:accuracy}
Over a sample set $\calD = \{(s^{(k)}, a^{(k)}, s'^{(k)})\}_{k=1}^n$:
\begin{equation}
\text{Acc}(\hat{T}; \calD) = \frac{1}{n} \sum_{k=1}^n \ind[\epsilon(s^{(k)}, a^{(k)}; \hat{T}) = 0].
\end{equation}
The agent's objective is to maximize $\text{Acc}(\hat{T}; \tau_B)$ subject to the budget $B$.
\end{definition}

\subsection{Challenges}
\label{sec:challenges}

Three challenges distinguish this setting from standard model learning:

\begin{enumerate}[nosep]
    \item \textbf{Tabula rasa semantics.}  No bijection between $\calA$ and semantic labels is known a priori; the mapping $a \mapsto \text{``meaning''}$ must be discovered from transitions.

    \item \textbf{Co-dependent representations.}  The object role assignment $\rho : \calO \to \calR$ depends on $T$ (we need to see how objects respond to actions to identify them as ``player'', ``wall'', etc.), while expressing $T$ compactly requires $\rho$.  This circularity connects to the frame problem~\cite{mccarthy1969some} in classical AI.

    \item \textbf{Test-time budget.}  The model must be learned online within $B$ interactions, precluding methods requiring thousands of gradient steps (Dreamer~\cite{hafner2020dreamer}) or extensive self-play (MuZero~\cite{schrittwieser2020muzero}).
\end{enumerate}


% ============================================================================
% SECTION 3: DUAL KNOWLEDGE REPRESENTATION
% ============================================================================

\section{Dual Knowledge Representation}
\label{sec:knowledge}

The agent maintains a \emph{knowledge state} $\calK_t$ at each timestep, consisting of two complementary representations that correspond to the "NL-JSON model" and "Python Symbolic Model" in the implementation.

\subsection{Action Knowledge Base}
\label{sec:action_kb}

\begin{definition}[Action Knowledge Base]
\label{def:action_kb}
The \textbf{action knowledge base} is $\calK^{\calA}_t = \{k_a\}_{a \in \calA}$, where each entry is:
\begin{equation}
k_a = (d_a, \calH_a, v_a, \sigma_a, \eta_a^{\text{no-eff}}),
\end{equation}
with components:
\begin{itemize}[nosep]
    \item $d_a \in \Sigma^*$: A natural language \textbf{definition} encoding the action's preconditions and effects over the object vocabulary, implicitly representing a STRIPS schema (Remark~\ref{rem:implicit_strips}).
    \item $\calH_a = \{(s^{(i)}, s'^{(i)}, \delta^{(i)}, \iota^{(i)})\}_{i=1}^{n_a}$: The \textbf{observation history}, where $\delta^{(i)}$ is the structured diff and $\iota^{(i)} \in \Sigma^*$ is the NL interpretation.
    \item $v_a \in \{0, 1\}$: A \textbf{verification flag}; $v_a = 1$ iff $d_a$ has been confirmed by $\kappa$ consecutive consistent observations (default $\kappa = 3$).
    \item $\sigma_a \in \bbN_0$: A \textbf{consistency counter} tracking consecutive observations consistent with $d_a$.
    \item $\eta_a^{\text{no-eff}} \in \bbN_0$: A \textbf{no-effect counter}; the action is marked \textbf{exhausted} when $\eta_a^{\text{no-eff}} \geq \kappa_{\text{exhaust}}$ (default $\kappa_{\text{exhaust}} = 8$).
\end{itemize}
\end{definition}

\begin{remark}[Connection to Safe Action Models]
\label{rem:sam}
Juba et al.~\cite{juba2021safe} define a \textbf{safe action model} $\hat{\alpha}$ as one where the learned preconditions are sufficient (but possibly stronger than necessary): $\pre(\hat{\alpha}) \supseteq \pre(\alpha)$, ensuring that the agent never attempts an action in an invalid state.  Our verification flag $v_a$ serves an analogous purpose: only actions with $v_a = 1$ are used in setup sequences, providing a safety guarantee that setup actions will produce their expected effects.
\end{remark}

\subsection{Environment Knowledge Base}
\label{sec:env_kb}

\begin{definition}[Environment Knowledge Base]
\label{def:env_kb}
The \textbf{environment knowledge base} is:
\begin{equation}
\calK^{\calM}_t = (\calO_t, \rho_t, \calC_t, \xi_t, \beta_t, \calI_t),
\end{equation}
where:
\begin{itemize}[nosep]
    \item $\calO_t = \{o_1, \ldots, o_{M_t}\}$: identified objects, each augmented with a type label $\ell_j \in \Sigma^*$ (``player'', ``wall'', ``goal'', etc.).
    \item $\rho_t : \calO_t \to \calR$: role assignment mapping objects to semantic roles $\calR = \{\texttt{agent}, \texttt{obstacle}, \texttt{pushable}, \texttt{goal}, \texttt{decoration}, \ldots\}$.  This defines the unary predicates of the logical state: $\mathsf{IsAgent}(o) \iff \rho_t(o) = \texttt{agent}$.
    \item $\calC_t \subset \Sigma^*$: natural language \textbf{movement constraints}, corresponding to the precondition components of our implicit STRIPS model.
    \item $\xi_t \in \Sigma^*$: \textbf{spatial structure} description (grid dimensions, borders, regions).
    \item $\beta_t \subset \Sigma^*$: \textbf{breakthrough discoveries}, mechanistic insights that fundamentally revise the model (e.g., ``unsupported objects fall due to gravity'').
    \item $\calI_t \subseteq \calH_\calL$: the current \textbf{logical state}, the set of ground atoms the agent believes to be true, obtained via lifting (Definition~\ref{def:lifting}).
\end{itemize}
\end{definition}

\begin{definition}[Full Knowledge State]
\label{def:full_knowledge}
The agent's knowledge state at time $t$ is $\calK_t = (\calK^{\calA}_t, \calK^{\calM}_t)$.  The initial state $\calK_0$ has empty definitions, histories, objects, and constraints.
\end{definition}

\begin{remark}[Dual Representation Design]
\label{rem:dual}
The dual representation, NL-JSON knowledge base plus executable Python model, reflects a fundamental tension in AI planning between \emph{declarative} and \emph{procedural} knowledge~\cite{ghallab2004automated}.  The NL-JSON component captures declarative knowledge (what the agent knows about the world), while the executable model provides procedural knowledge (how to predict the next state).  This parallels the ``thinking fast and slow'' architecture~\cite{kahneman2011thinking}: the VLM provides slow, deliberative reasoning while the executable model enables fast, automatic prediction.
\end{remark}


% ============================================================================
% SECTION 4: VLM ORACLE AND TRANSITION ANALYSIS
% ============================================================================

\section{VLM Oracle and Transition Analysis}
\label{sec:vlm_oracle}

The VLM serves as an oracle for transition interpretation, analogous to the ``teacher'' in the Exact Learning Model~\cite{angluin1987learning} but operating over visual-linguistic inputs.

\subsection{Diff Computation}
\label{sec:diff}

Before invoking the VLM, a structured diff is computed between consecutive states.

\begin{definition}[State Diff]
\label{def:diff}
The \textbf{state diff} between $s_t$ and $s_{t+1}$ is:
\begin{equation}
\Delta_t = (\calX_t^{\text{pixel}}, \calX_t^{\text{obj}}, r_t),
\end{equation}
where:
\begin{itemize}[nosep]
    \item $\calX_t^{\text{pixel}} = \{(i, j, s_t[i,j], s_{t+1}[i,j]) : s_t[i,j] \neq s_{t+1}[i,j]\}$: pixel-level changes.
    \item $\calX_t^{\text{obj}} \subseteq \calO \times \{\texttt{appeared}, \texttt{disappeared}, \texttt{moved}, \texttt{reshaped}, \texttt{recolored}\}$: object-level changes computed by comparing $\Omega(s_t)$ and $\Omega(s_{t+1})$ via a three-pass matching algorithm (color match $\to$ spatial overlap $\to$ nearest-neighbor).
    \item $r_t \in \Sigma^*$: a text summary (e.g., ``Color 2 block moved from $(24,12)$ to $(20,12)$'').
\end{itemize}
\end{definition}

\begin{remark}[Relation to Contrastive Methods]
The diff computation is conceptually related to the contrastive losses used in C-SWM~\cite{kipf2020contrastive}: both identify what changed between states.  C-SWM uses a hinge-loss in embedding space; our diff operates in pixel/object space and produces a symbolic description suitable for VLM consumption.
\end{remark}

\subsection{Action Analysis (VLM Call 1)}
\label{sec:action_analysis}

\begin{definition}[Action Analysis Function]
\label{def:action_analysis}
The \textbf{action analysis function} is:
\begin{equation}
f_{\text{act}} : \calS \times \calS \times \calA \times \Delta \times \calK_t \longrightarrow (\iota, d'_a, \gamma, \eta),
\end{equation}
where $\iota \in \Sigma^*$ is a transition interpretation, $d'_a \in \Sigma^* \cup \{\bot\}$ is an updated action definition (or $\bot$ for no change), $\gamma \in \{\texttt{true}, \texttt{false}, \bot\}$ is a consistency judgment, and $\eta \in \{0, 1\}$ is an effect flag.
\end{definition}

This function is realized by prompting the VLM with the before/after grids (rendered as ASCII art), the computed diff, and the current knowledge state.  The prompt structure follows the \emph{chain-of-thought} paradigm~\cite{wei2022chain}, requiring the VLM to first describe what it observes, then compare with the current definition, and finally produce a judgment.

\begin{remark}[Relation to LLM-Based Planning]
The use of VLM prompts for action analysis connects to several lines of work:
\begin{itemize}[nosep]
    \item \textbf{LLMs as Zero-Shot Planners}~\cite{huang2022language}: using LLM knowledge to decompose tasks into action sequences.  Our VLM similarly leverages pre-trained knowledge to interpret novel game mechanics.
    \item \textbf{SayCan}~\cite{ahn2022can}: grounding LLM plans in affordances.  Our diff computation provides the ``grounding'' signal that constrains VLM interpretations to be consistent with observed state changes.
    \item \textbf{ProgPrompt}~\cite{singh2023progprompt}: LLM-generated programs for robotics.  Our compilation step (Section~\ref{sec:compilation}) similarly generates executable code from NL specifications.
\end{itemize}
\end{remark}

\subsection{Environment Analysis (VLM Call 2)}
\label{sec:env_analysis}

\begin{definition}[Environment Analysis Function]
\label{def:env_analysis}
The \textbf{environment analysis function} is:
\begin{equation}
f_{\text{env}} : \calS \times \Delta \times \calK_t \longrightarrow (\calO'_t, \rho'_t, \calC'_t, \xi'_t, \beta'_t),
\end{equation}
producing updated environment knowledge components.  This function identifies new objects, refines role assignments, and discovers environmental constraints.
\end{definition}

The environment analysis performs \emph{predicate grounding} in the sense of Athalye et al.~\cite{athalye2024pixels}: it maps raw visual observations to logical predicates.  Unlike their approach, which uses a fixed set of PDDL predicates, our system discovers predicates dynamically as the agent encounters new phenomena.

\subsection{Knowledge Update Rule}
\label{sec:update}

\begin{definition}[Knowledge Update]
\label{def:update}
Given analysis outputs, the knowledge state is updated by:
\begin{align}
d_{a_t} &\leftarrow \begin{cases} d'_{a_t} & \text{if } d'_{a_t} \neq \bot \\ d_{a_t} & \text{otherwise} \end{cases} \label{eq:def_update} \\
\calH_{a_t} &\leftarrow \calH_{a_t} \cup \{(s_t, s_{t+1}, \delta_t, \iota_t)\} \label{eq:history_update} \\
\sigma_{a_t} &\leftarrow \begin{cases} \sigma_{a_t} + 1 & \text{if } \gamma = \texttt{true} \\ 0 & \text{if } \gamma = \texttt{false} \end{cases} \label{eq:consistency_update} \\
v_{a_t} &\leftarrow \ind[\sigma_{a_t} \geq \kappa] \label{eq:verify_update} \\
\calO_{t+1} &\leftarrow \textsc{Merge}(\calO_t, \calO'_t) \label{eq:obj_merge} \\
\calC_{t+1} &\leftarrow \calC_t \cup \calC'_t \label{eq:constraint_merge} \\
\beta_{t+1} &\leftarrow \beta_t \cup \beta'_t \label{eq:breakthrough_merge}
\end{align}
\end{definition}

\begin{remark}[Connection to OLAM]
\label{rem:olam}
Lamanna et al.~\cite{lamanna2021olam} formalize \textbf{Online Learning of Action Models} (OLAM), defining a planning domain $M = \langle P, O, H \rangle$ (predicates, operator names, action schemas) with its finite-state machine $M(C) = \langle S, A, \delta \rangle$.  Their framework requires two properties:
\begin{enumerate}[nosep]
    \item \textbf{Correctness} (Theorem~2 in~\cite{lamanna2021olam}): $M(C)$ correctly approximates the ground truth $M'(C)$ from $s_0$ if for every reachable state $s_n$ and action $a$: $\langle s_n, a, s' \rangle \in \delta'$ implies $\langle s_n, a, s \rangle \in \delta$ for some $s' \supseteq s$.
    \item \textbf{Integrity} (Theorem~4 in~\cite{lamanna2021olam}): $M(C)$ integrally approximates $M'(C)$ if reachable transitions in $M'$ are eventually representable in $M$.
\end{enumerate}
Our verification mechanism (Eqs.~\ref{eq:consistency_update}--\ref{eq:verify_update}) achieves analogous properties: the consistency counter ensures correctness ($\kappa$ consecutive confirmations), while continued exploration ensures integrity (unverified actions are prioritized for testing).  Notably, OLAM is the only classical action model learning algorithm that addresses \emph{active exploration}, it constructs goal formulas to reach states where executing specific actions yields maximal information, closely paralleling our setup sequence mechanism (Definition~\ref{def:explore_policy}).  The SAM family~\cite{juba2021safe, le2024pisam, mordoch2024csam} provides the safety guarantee that learned models never produce plans that would fail in the real environment.
\end{remark}


% ============================================================================
% SECTION 5: EXECUTABLE WORLD MODEL AND COMPILATION
% ============================================================================

\section{Executable World Model and Compilation Pipeline}
\label{sec:executable_model}

\subsection{Executable World Model}
\label{sec:exec_model_def}

\begin{definition}[Executable World Model]
\label{def:exec_model}
An \textbf{executable world model} is a deterministic function $\hat{T}_\theta : \calS \times \calA \to \calS$ parameterized by a Python program $\theta \in \Sigma^*$, structured as:
\begin{equation}
\label{eq:model_structure}
\hat{T}_\theta(s, a) = (\calC^{\text{post}}_\theta \circ \calP_\theta \circ \calT_{a,\theta} \circ \calC^{\text{pre}}_\theta)(s),
\end{equation}
where:
\begin{itemize}[nosep]
    \item $\calC^{\text{pre}}_\theta, \calC^{\text{post}}_\theta : \calS \to \calS$: pre- and post-transition \textbf{constraint enforcement} (boundary checks, budget limits).  If $\calC^{\text{pre}}_\theta(s) = s$ with an ``action blocked'' flag, the action is a no-op.
    \item $\calT_{a,\theta} : \calS \to \calS$: the \textbf{action-specific transition}, implementing the STRIPS-like semantics of action $a$.
    \item $\calP_\theta : \calS \to \calS$: \textbf{physics rules} (action-independent effects such as gravity, cascading, etc.).
\end{itemize}
\end{definition}

This decomposition mirrors the structure of PDDL domains~\cite{fox2003pddl2}, where action effects and domain axioms are specified separately.  It also reflects the BaseGame architecture in our implementation, where \texttt{handle\_action} implements $\calT_{a,\theta}$, \texttt{update} implements $\calP_\theta$, and boundary checks implement $\calC_\theta$.

\begin{proposition}[Executable Model as STRIPS Realization]
\label{prop:strips_realization}
If the executable model $\hat{T}_\theta$ is correct (i.e., $\hat{T}_\theta(s, a) = T(s, a)$ for all visited $(s, a)$), and the environment is STRIPS-representable with schemas $\{\alpha_a\}$ under lifting $\Lambda$, then the action-specific transition $\calT_{a,\theta}$ implements the grounded STRIPS transition:
\begin{equation}
\Lambda(\calT_{a,\theta}(s)) = (\Lambda(s) \setminus \eff^-(\alpha_a)) \cup \eff^+(\alpha_a).
\end{equation}
The Python program $\theta$ thus serves as a procedural encoding of the learned STRIPS schemas, with the constraint functions $\calC_\theta$ encoding preconditions and the physics function $\calP_\theta$ encoding domain axioms (conditional effects that are action-independent).
\end{proposition}

\subsection{Compilation: NL-JSON $\to$ Executable Model}
\label{sec:compilation}

\begin{definition}[Model Compilation Function]
\label{def:compile}
The \textbf{compilation function} is:
\begin{equation}
f_{\text{compile}} : \calK^{\calA}_t \times \calK^{\calM}_t \times \Sigma^*_{\text{API}} \times (\theta_{\text{prev}}, E_{\text{prev}}) \longrightarrow \theta \in \Sigma^*,
\end{equation}
where $\Sigma^*_{\text{API}}$ is the target API specification (base classes, method signatures), and $(\theta_{\text{prev}}, E_{\text{prev}})$ is the previous program and its error report.
\end{definition}

This compilation step is the central bridge between declarative (NL) and procedural (code) knowledge.  It is closely related to:

\begin{itemize}[nosep]
    \item \textbf{Code as Policies}~\cite{liang2023code}: using LLMs to generate robot policy code from NL specifications.  Our compilation similarly generates a Python program, but targets world model prediction rather than action selection.
    \item \textbf{LLM+P}~\cite{liu2023llmp}: translating NL problem descriptions into PDDL for classical planners.  Our compilation translates NL action definitions into executable Python.
    \item \textbf{DreamCoder}~\cite{ellis2021dreamcoder}: synthesizing programs from examples via a library of learned abstractions.  Our system builds toward a similar compositional program structure, but uses VLM-guided synthesis rather than enumerative search.
\end{itemize}

\subsubsection{Compilation Trigger Conditions}
\label{sec:compile_trigger}

Compilation is triggered when:
\begin{enumerate}[nosep]
    \item No executable model exists yet and $|\tau_t| \geq n_{\min}$ (sufficient observations gathered);
    \item The previous prediction was incorrect ($\epsilon_t > 0$);
    \item At least $\Delta_{\text{compile}}$ actions have elapsed since the last compilation.
\end{enumerate}

\subsection{Self-Refinement Loop}
\label{sec:self_refine}

The compilation process implements a \textbf{Self-Refine}~\cite{madaan2024self} loop: the VLM generates an initial program, receives feedback (error signal), and iteratively improves it.

\begin{definition}[Self-Refine Iteration]
\label{def:self_refine}
At each compilation step $k$, the refinement follows:
\begin{align}
\theta^{(0)} &= f_{\text{compile}}(\calK_t, \Sigma^*_{\text{API}}, \bot) & \text{(initial generation)} \\
E^{(k)} &= \textsc{Evaluate}(\hat{T}_{\theta^{(k)}}, \tau_t) & \text{(feedback)} \\
\theta^{(k+1)} &= f_{\text{compile}}(\calK_t, \Sigma^*_{\text{API}}, (\theta^{(k)}, E^{(k)})) & \text{(refinement)}
\end{align}
until $E^{(k)} = \varnothing$ (no errors) or $k = k_{\max}$ (maximum iterations).
\end{definition}

This three-phase loop (generate $\to$ evaluate $\to$ refine) exactly mirrors the Self-Refine framework, and is related to Reflexion~\cite{shinn2024reflexion}, where verbal self-reflection replaces scalar rewards as the learning signal.  The insight from Reflexion, that natural language feedback is more informative than scalar rewards, directly motivates our structured error signals.

\begin{remark}[Relation to Voyager's Skill Library]
\label{rem:voyager}
Voyager~\cite{wang2024voyager} maintains a skill library of verified code snippets that are composed for complex tasks.  Our executable model $\hat{T}_\theta$ can be viewed as a structured skill library where each action-specific transition $\calT_{a,\theta}$ is a verified ``skill'' that is composed with physics rules $\calP_\theta$ and constraints $\calC_\theta$ to form the complete world model.
\end{remark}


% ============================================================================
% SECTION 6: VERIFICATION AND ERROR DIAGNOSIS
% ============================================================================

\section{Verification and Error Diagnosis}
\label{sec:verification}

\subsection{Prediction and Per-Transition Verification}
\label{sec:predict_verify}

Given $\hat{T}_\theta$, each new transition $(s_t, a_t, s_{t+1})$ is verified:
\begin{equation}
\hat{s}_{t+1} = \hat{T}_\theta(s_t, a_t), \qquad \epsilon_t = \epsilon(s_t, a_t; \hat{T}_\theta).
\end{equation}
The prediction is \textbf{correct} iff $\hat{s}_{t+1} = s_{t+1}$ (equivalently, $\epsilon_t = 0$).

\subsection{Regression Testing}
\label{sec:regression}

After each recompilation $\theta \to \theta'$, the agent performs regression testing:
\begin{equation}
\text{Acc}(\hat{T}_{\theta'}; \tau_t) = \frac{1}{t} \sum_{k=0}^{t-1} \ind[\hat{T}_{\theta'}(s_k, a_k) = s_{k+1}].
\end{equation}
A \textbf{regression} is a transition correctly predicted by $\theta$ but not by $\theta'$:
\begin{equation}
\calR_{\theta \to \theta'} = \{k : \hat{T}_\theta(s_k, a_k) = s_{k+1} \;\wedge\; \hat{T}_{\theta'}(s_k, a_k) \neq s_{k+1}\}.
\end{equation}
Regressions receive highest priority in the error signal for the next compilation.

\subsection{Structured Error Signal}
\label{sec:error_signal}

\begin{definition}[Error Signal]
\label{def:error_signal}
When $\epsilon_t > 0$, the agent constructs:
\begin{equation}
E_t = (\calX_t^{\text{err}}, \ell_t, d_{a_t}, \calC_t, \calR_t),
\end{equation}
where:
\begin{itemize}[nosep]
    \item $\calX_t^{\text{err}} = \{(i,j, \hat{s}_{t+1}[i,j], s_{t+1}[i,j]) : \hat{s}_{t+1}[i,j] \neq s_{t+1}[i,j]\}$: cells where prediction disagrees with reality.
    \item $\ell_t \in \Sigma^*$: execution log recording which model components fired.
    \item $d_{a_t}$: current NL definition of the action.
    \item $\calC_t$: current constraints.
    \item $\calR_t$: regression set (if recompilation occurred).
\end{itemize}
\end{definition}

This cell-level, mechanistically-attributed error signal is far richer than the undifferentiated gradient signals of standard MBRL~\cite{hafner2020dreamer}.  It enables targeted debugging analogous to the ``error localization'' step in program synthesis~\cite{gulwani2017program}.

\subsection{Safety and PAC-Style Guarantees}
\label{sec:pac_guarantees}

We connect our verification mechanism to the formal safety framework of action model learning~\cite{juba2021safe, le2024pisam, mordoch2024csam}.

\begin{definition}[Safe Action Model ,  adapted from Juba et al.~\cite{juba2021safe}]
\label{def:safe_action_model}
An action model $M'$ with preconditions $\pre_{M'}$ and effects $\eff_{M'}$ is \textbf{safe} with respect to the true action model $M$ iff for every state $s$ and ground action $a$:
\begin{equation}
\pre_{M'}(a) \subseteq s \;\implies\; \bigl(\pre_M(a) \subseteq s \;\wedge\; a_{M'}(s) = a_M(s)\bigr).
\end{equation}
Informally, whenever the learned model deems an action applicable, the true model also deems it applicable and the effects agree.  This ensures that plans generated using $M'$ are \emph{executable} under the true model $M$.
\end{definition}

\begin{definition}[$\epsilon$-Adequacy ,  adapted from Juba et al.~\cite{juba2021safe}]
\label{def:adequate}
An action model $M$ is \textbf{$\epsilon$-adequate} if, with probability at most $\epsilon$, a trajectory $T$ drawn from the problem distribution contains a transition $\langle s, a, s' \rangle$ where $s$ does not satisfy $\pre_M(a)$ or there exists a literal in $s' \setminus s$ not accounted for by $\eff_M(a)$.
\end{definition}

\begin{proposition}[Verification as PAC Condition]
\label{prop:pac_verification}
Let $\hat{\alpha}_a = (\widehat{\pre}(a), \widehat{\eff}^+(a), \widehat{\eff}^-(a))$ be the implicit STRIPS schema extracted from the verified definition $d_a$ (with $v_a = 1$).  If the VLM's consistency judgments are correct (i.e., $\gamma = \texttt{true}$ implies the observation is genuinely consistent with $d_a$), then after $\kappa$ consecutive confirmations, the probability that $\hat{\alpha}_a$ misclassifies an observable transition is bounded by:
\begin{equation}
\Pr[\hat{\alpha}_a \text{ is incorrect on a random } s \sim \mu] \leq (1 - p_{\min})^\kappa,
\end{equation}
where $p_{\min} = \min_{s \in \text{supp}(\mu)} \mu(s)$ and $\mu$ is the distribution over states visited during exploration.  For uniform exploration over $m$ distinct states, this gives $(1 - 1/m)^\kappa$.
\end{proposition}

\begin{proof}[Proof sketch]
Each consistency confirmation is an independent test of $\hat{\alpha}_a$ on a state drawn from the exploration distribution.  If $\hat{\alpha}_a$ were incorrect on a state $s^*$ with probability $\mu(s^*) \geq p_{\min}$, the probability of $\kappa$ consecutive confirmations without encountering $s^*$ is at most $(1 - p_{\min})^\kappa$.
\end{proof}

\begin{remark}[Connection to SAM Safety Hierarchy]
Our verification mechanism provides safety guarantees at multiple levels:
\begin{itemize}[nosep]
    \item \textbf{Setup-sequence safety}: only actions with $v_a = 1$ are used in setup sequences (Definition~\ref{def:explore_policy}), analogous to SAM's guarantee that plans using safe action models are executable~\cite{juba2021safe}.
    \item \textbf{Compilation safety}: the regression testing mechanism (Section~\ref{sec:regression}) ensures that the executable model does not regress on previously correct predictions, analogous to OLAM's correctness guarantee~\cite{lamanna2021olam}.
    \item \textbf{Adequacy convergence}: as $|\calH_a|$ grows, the VLM observes more diverse contexts, making the implicit action model $\epsilon$-adequate (Definition~\ref{def:adequate}) with high probability.  PI-SAM's sample complexity bound~\cite{le2024pisam}, $O(\frac{1}{\eta^2}(|\calA||\calF| + \log(1/\delta)))$ under $\eta$-bounded concealment, provides a theoretical upper bound on the number of observations needed.
\end{itemize}
\end{remark}


% ============================================================================
% SECTION 7: ACTIVE EXPLORATION
% ============================================================================

\section{Active Exploration}
\label{sec:exploration}

\subsection{Exploration Policy}
\label{sec:explore_policy}

\begin{definition}[Exploration Policy]
\label{def:explore_policy}
The \textbf{exploration policy} is:
\begin{equation}
\pi_{\text{explore}} : \calS \times \calK_t \longrightarrow (a_{\text{target}}, \mathbf{a}_{\text{setup}}),
\end{equation}
where $a_{\text{target}} \in \calA$ is the action to investigate and $\mathbf{a}_{\text{setup}} = (a'_1, \ldots, a'_m) \in \calA^m$ is a \textbf{setup sequence} of verified actions executed to reach an informative state.
\end{definition}

The setup sequence concept connects to the macro-action and option framework~\cite{sutton1999between}: each setup sequence is a temporally extended action that navigates to a state where the target action is informative.  Only actions with $v_a = 1$ are used in setup sequences, providing the safety guarantee of Remark~\ref{rem:sam}.

\subsection{Information-Theoretic Motivation}
\label{sec:info_gain}

The exploration policy is motivated by maximizing information gain about the transition function.  Following VIME~\cite{houthooft2016vime}, we define:

\begin{definition}[Information Gain]
\label{def:info_gain}
The \textbf{information gain} from executing action $a$ in state $s$ is:
\begin{equation}
\calI(s, a; \calK_t) = H[\hat{T}(\cdot \mid s, a; \calK_t)] - H[\hat{T}(\cdot \mid s, a; \calK_{t+1})],
\end{equation}
where $H[\cdot]$ denotes the entropy of the predictive distribution over outcomes.  In the deterministic case, this reduces to a binary signal: $\calI(s, a; \calK_t) > 0$ iff the observation resolves genuine uncertainty about $T(s, a)$.
\end{definition}

In practice, the VLM-based exploration policy approximates information gain through a priority ordering:

\begin{enumerate}[nosep]
    \item \textbf{Untested actions}: $|\calH_a| = 0$ (maximum uncertainty).
    \item \textbf{Unverified actions}: $v_a = 0,\; \sigma_a < \kappa$ (partially understood).
    \item \textbf{Context diversity}: for verified actions, seek novel states not represented in $\calH_a$.
    \item \textbf{Hypothesis testing}: actively test edge cases of current definitions.
\end{enumerate}

\begin{remark}[Connection to Exploration Methods]
This priority ordering has connections to several exploration paradigms:
\begin{itemize}[nosep]
    \item \textbf{R-MAX}~\cite{brafman2002rmax}: the ``untested/unverified'' distinction mirrors R-MAX's known/unknown partition (Definition~\ref{def:known_unknown}).
    \item \textbf{ICM}~\cite{pathak2017curiosity}: intrinsic curiosity driven by prediction error.  Our system prioritizes actions/states where the model is uncertain, analogous to ICM's prediction-error bonus.
    \item \textbf{Plan2Explore}~\cite{sekar2020planning}: ensemble disagreement for exploration in model-based RL.  While we lack an ensemble, the consistency counter $\sigma_a$ measures a form of ``agreement'' across observations.
    \item \textbf{Go-Explore}~\cite{ecoffet2021first}: maintaining an archive of ``interesting'' states and returning to them.  Our setup sequences enable systematic return to specific states.
\end{itemize}
\end{remark}

\subsection{Hard Guard Mechanism}
\label{sec:guard}

A deterministic guard prevents degenerate exploration:
\begin{equation}
\label{eq:guard}
a \notin \calA_{\text{allowed}}(s_t) \quad \text{if} \quad s_t = s_{t-1} \;\wedge\; a = a_{t-1} \;\wedge\; \eta_{t-1} = 0.
\end{equation}
The guard is lifted when a different action changes the state, ensuring $\calA_{\text{allowed}}(s_t) \neq \varnothing$ as long as at least one action produces an effect in some reachable state.


% ============================================================================
% SECTION 8: CO-EVOLUTION AND CONVERGENCE
% ============================================================================

\section{Co-Evolution of Object and Action Knowledge}
\label{sec:coevolution}

A defining feature of our approach is that object representations and action definitions are \emph{mutually dependent} and evolve together, forming a \emph{coupled dynamical system} over the knowledge state.

\subsection{Coupled Update Dynamics}
\label{sec:coupled}

At each timestep, the environment analysis refines objects and roles $(\calO_t, \rho_t)$ based on action transitions, while action analysis refines definitions $\{d_a\}$ using the current object vocabulary:
\begin{align}
(\calO_{t+1}, \rho_{t+1}) &= g_{\calO}(\calO_t, \rho_t, \{(s_k, a_k, s_{k+1})\}_{k \leq t}) \label{eq:obj_update} \\
d_{a, t+1} &= g_d(d_{a,t}, \calO_{t+1}, \rho_{t+1}, \calH_{a, \leq t}) \label{eq:def_update_coupled}
\end{align}
where $g_{\calO}$ is realized by $f_{\text{env}}$ and $g_d$ by $f_{\text{act}}$.  The updated object model (Eq.~\ref{eq:obj_update}) feeds into the action definition update (Eq.~\ref{eq:def_update_coupled}), and vice versa.

This co-evolution is related to:
\begin{itemize}[nosep]
    \item \textbf{EM algorithms}: alternating between updating the ``model'' (E-step $\approx$ action analysis) and the ``parameters'' (M-step $\approx$ object roles).
    \item \textbf{Abstraction-refinement loops} in CEGAR~\cite{clarke2003counterexample}: iteratively refining an abstract model when counterexamples are found.
    \item \textbf{DreamerV3's} ``learn the world, then plan''~\cite{hafner2023dreamerv3}: alternating between world model learning and policy optimization.
\end{itemize}

\subsection{Convergence}
\label{sec:convergence}

% ============================================================================
% BIBLIOGRAPHY
% ============================================================================

\bibliographystyle{plain}

\begin{thebibliography}{50}

\bibitem{ahn2022can}
M.~Ahn, A.~Brohan, N.~Brown, et~al.
\newblock Do As I Can, Not As I Say: Grounding Language in Robotic Affordances.
\newblock In \emph{CoRL}, 2022.

\bibitem{angluin1987learning}
D.~Angluin.
\newblock Learning Regular Sets from Queries and Counterexamples.
\newblock \emph{Information and Computation}, 75(2):87--106, 1987.

\bibitem{arora2018review}
A.~Arora, H.~Fiorino, D.~Pellier, M.~M{\'e}tivier, and S.~Pesty.
\newblock A Review of Learning Planning Action Models.
\newblock \emph{The Knowledge Engineering Review}, 33:e20, 2018.

\bibitem{athalye2024pixels}
A.~Athalye, N.~Mandi, T.~Silver, and Y.~Liang.
\newblock Pixels to Predicates: Learning Symbolic World Models from Vision.
\newblock \emph{arXiv preprint}, 2024.

\bibitem{bellemare2016unifying}
M.~Bellemare, S.~Srinivasan, G.~Ostrovski, T.~Schaul, D.~Saxton, and R.~Munos.
\newblock Unifying Count-Based Exploration and Intrinsic Motivation.
\newblock In \emph{NeurIPS}, 2016.

\bibitem{brafman2002rmax}
R.~I.~Brafman and M.~Tennenholtz.
\newblock R-MAX, A General Polynomial Time Algorithm for Near-Optimal Reinforcement Learning.
\newblock \emph{JMLR}, 3:213--231, 2002.

\bibitem{chollet2019measure}
F.~Chollet.
\newblock On the Measure of Intelligence.
\newblock \emph{arXiv:1911.01547}, 2019.

\bibitem{clarke2003counterexample}
E.~Clarke, O.~Grumberg, S.~Jha, Y.~Lu, and H.~Veith.
\newblock Counterexample-Guided Abstraction Refinement for Symbolic Model Checking.
\newblock \emph{JACM}, 50(5):752--794, 2003.

\bibitem{cresswell2011acquiring}
S.~N.~Cresswell, T.~L.~McCluskey, and M.~M.~West.
\newblock Acquiring Planning Domain Models Using LOCM.
\newblock \emph{The Knowledge Engineering Review}, 28(2):195--213, 2013.

\bibitem{ecoffet2021first}
A.~Ecoffet, J.~Huizinga, J.~Lehman, K.~O.~Stanley, and J.~Clune.
\newblock First Return, Then Explore.
\newblock \emph{Nature}, 590:580--586, 2021.

\bibitem{ellis2021dreamcoder}
K.~Ellis, C.~Wong, M.~Nye, et~al.
\newblock DreamCoder: Bootstrapping Inductive Program Synthesis with Wake-Sleep Library Learning.
\newblock In \emph{PLDI}, 2021.

\bibitem{fikes1971strips}
R.~E.~Fikes and N.~J.~Nilsson.
\newblock STRIPS: A New Approach to the Application of Theorem Proving to Problem Solving.
\newblock \emph{Artificial Intelligence}, 2(3--4):189--208, 1971.

\bibitem{fox2003pddl2}
M.~Fox and D.~Long.
\newblock PDDL2.1: An Extension to PDDL for Expressing Temporal Planning Domains.
\newblock \emph{JAIR}, 20:61--124, 2003.

\bibitem{ghallab2004automated}
M.~Ghallab, D.~Nau, and P.~Traverso.
\newblock \emph{Automated Planning: Theory and Practice}.
\newblock Morgan Kaufmann, 2004.

\bibitem{givan2003equivalence}
R.~Givan, T.~Dean, and M.~Greig.
\newblock Equivalence Notions and Model Minimization in Markov Decision Processes.
\newblock \emph{Artificial Intelligence}, 147(1--2):163--223, 2003.

\bibitem{gulwani2017program}
S.~Gulwani, O.~Polozov, and R.~Singh.
\newblock Program Synthesis.
\newblock \emph{Foundations and Trends in Programming Languages}, 4(1--2):1--119, 2017.

\bibitem{ha2018world}
D.~Ha and J.~Schmidhuber.
\newblock World Models.
\newblock \emph{arXiv:1803.10122}, 2018.

\bibitem{hafner2020dreamer}
D.~Hafner, T.~Lillicrap, J.~Ba, and M.~Norouzi.
\newblock Dream to Control: Learning Behaviors by Latent Imagination.
\newblock In \emph{ICLR}, 2020.

\bibitem{hafner2023dreamerv3}
D.~Hafner, J.~Pasukonis, J.~Ba, and T.~Lillicrap.
\newblock Mastering Diverse Domains through World Models.
\newblock \emph{arXiv:2301.04104}, 2023.

\bibitem{he2017connected}
L.~He, X.~Ren, Q.~Gao, X.~Zhao, B.~Yao, and Y.~Chao.
\newblock The Connected-Component Labeling Problem: A Review of State-of-the-Art Algorithms.
\newblock \emph{Pattern Recognition}, 70:25--43, 2017.

\bibitem{houthooft2016vime}
R.~Houthooft, X.~Chen, Y.~Duan, J.~Schulman, F.~De~Turck, and P.~Abbeel.
\newblock VIME: Variational Information Maximizing Exploration.
\newblock In \emph{NeurIPS}, 2016.

\bibitem{huang2022language}
W.~Huang, P.~Abbeel, D.~Pathak, and I.~Mordatch.
\newblock Language Models as Zero-Shot Planners: Extracting Actionable Knowledge for Embodied Agents.
\newblock In \emph{ICML}, 2022.

\bibitem{juba2021safe}
B.~Juba and R.~S.~Stern.
\newblock Safe Action Model Learning.
\newblock \emph{arXiv:2107.09336}, 2021.

\bibitem{kahneman2011thinking}
D.~Kahneman.
\newblock \emph{Thinking, Fast and Slow}.
\newblock Farrar, Straus and Giroux, 2011.

\bibitem{kakade2003sample}
S.~M.~Kakade.
\newblock On the Sample Complexity of Reinforcement Learning.
\newblock PhD thesis, University College London, 2003.

\bibitem{kipf2020contrastive}
T.~Kipf, E.~van~der~Pol, and M.~Welling.
\newblock Contrastive Learning of Structured World Models.
\newblock In \emph{ICLR}, 2020.

\bibitem{konidaris2018skills}
G.~Konidaris, L.~P.~Kaelbling, and T.~Lozano-P{\'e}rez.
\newblock From Skills to Symbols: Learning Symbolic Representations for Abstract High-Level Planning.
\newblock \emph{JAIR}, 61:215--289, 2018.

\bibitem{li2006towards}
L.~Li, T.~J.~Walsh, and M.~L.~Littman.
\newblock Towards a Unified Theory of State Abstraction for MDPs.
\newblock In \emph{ISAIM}, 2006.

\bibitem{liang2023code}
J.~Liang, W.~Huang, F.~Xia, P.~Xu, K.~Hausman, B.~Ichter, P.~Florence, and A.~Zeng.
\newblock Code as Policies: Language Model Programs for Embodied Control.
\newblock In \emph{ICRA}, 2023.

\bibitem{liu2023llmp}
B.~Liu, Y.~Jiang, X.~Zhang, Q.~Liu, S.~Zhang, J.~Biswas, and P.~Stone.
\newblock LLM+P: Empowering Large Language Models with Optimal Planning Proficiency.
\newblock \emph{arXiv:2304.11477}, 2023.

\bibitem{locatello2020object}
F.~Locatello, D.~Weissenborn, T.~Unterthiner, et~al.
\newblock Object-Centric Learning with Slot Attention.
\newblock In \emph{NeurIPS}, 2020.

\bibitem{madaan2024self}
A.~Madaan, N.~Tandon, P.~Gupta, et~al.
\newblock Self-Refine: Iterative Refinement with Self-Feedback.
\newblock In \emph{NeurIPS}, 2024.

\bibitem{mccarthy1969some}
J.~McCarthy and P.~J.~Hayes.
\newblock Some Philosophical Problems from the Standpoint of Artificial Intelligence.
\newblock \emph{Machine Intelligence}, 4:463--502, 1969.

\bibitem{mcdermott1998pddl}
D.~McDermott, M.~Ghallab, A.~Howe, C.~Knoblock, A.~Ram, M.~Veloso, D.~Weld, and D.~Wilkins.
\newblock PDDL, The Planning Domain Definition Language.
\newblock Technical Report CVC TR-98-003, Yale Center for Computational Vision and Control, 1998.

\bibitem{nottingham2023embodied}
K.~Nottingham, P.~Ammanabrolu, A.~Suhr, Y.~Choi, H.~Hajishirzi, S.~Singh, and Y.~Bisk.
\newblock Do Embodied Agents Dream of Pixelated Sheep: Embodied Decision Making using Language Guided World Modelling.
\newblock In \emph{ICML}, 2023.

\bibitem{pathak2017curiosity}
D.~Pathak, P.~Agrawal, A.~A.~Efros, and T.~Darrell.
\newblock Curiosity-Driven Exploration by Self-Supervised Prediction.
\newblock In \emph{ICML}, 2017.

\bibitem{santoro2017simple}
A.~Santoro, D.~Raposo, D.~G.~T.~Barrett, et~al.
\newblock A Simple Neural Network Module for Relational Reasoning.
\newblock In \emph{NeurIPS}, 2017.

\bibitem{schrittwieser2020muzero}
J.~Schrittwieser, I.~Antonoglou, T.~Hubert, et~al.
\newblock Mastering Atari, Go, Chess and Shogi by Planning with a Learned Model.
\newblock \emph{Nature}, 588:604--609, 2020.

\bibitem{sekar2020planning}
R.~Sekar, O.~Rybkin, K.~Daniilidis, P.~Abbeel, D.~Hafner, and D.~Pathak.
\newblock Planning to Explore via Self-Supervised World Models.
\newblock In \emph{ICML}, 2020.

\bibitem{shinn2024reflexion}
N.~Shinn, F.~Cassano, A.~Gopinath, K.~R.~Narasimhan, and S.~Yao.
\newblock Reflexion: Language Agents with Verbal Reinforcement Learning.
\newblock In \emph{NeurIPS}, 2024.

\bibitem{silver2018general}
D.~Silver, T.~Hubert, J.~Schrittwieser, et~al.
\newblock A General Reinforcement Learning Algorithm that Masters Chess, Shogi, and Go through Self-Play.
\newblock \emph{Science}, 362(6419):1140--1144, 2018.

\bibitem{singh2023progprompt}
I.~Singh, V.~Blukis, A.~Mousavian, A.~Goyal, D.~Xu, J.~Tremblay, D.~Fox, J.~Thomason, and A.~Garg.
\newblock ProgPrompt: Generating Situated Robot Task Plans using Large Language Models.
\newblock In \emph{ICRA}, 2023.

\bibitem{stern2017efficient}
R.~Stern and B.~Juba.
\newblock Efficient, Safe, and Probably Approximately Complete Learning of Action Models.
\newblock In \emph{IJCAI}, 2017.

\bibitem{strehl2009reinforcement}
A.~L.~Strehl, L.~Li, and M.~L.~Littman.
\newblock Reinforcement Learning in Finite MDPs: PAC Analysis.
\newblock \emph{JMLR}, 10:2413--2444, 2009.

\bibitem{sutton1991dyna}
R.~S.~Sutton.
\newblock Dyna, An Integrated Architecture for Learning, Planning, and Reacting.
\newblock \emph{SIGART Bulletin}, 2(4):160--163, 1991.

\bibitem{sutton1999between}
R.~S.~Sutton, D.~Precup, and S.~Singh.
\newblock Between MDPs and Semi-MDPs: A Framework for Temporal Abstraction in Reinforcement Learning.
\newblock \emph{Artificial Intelligence}, 112(1--2):181--211, 1999.

\bibitem{veerapaneni2020entity}
R.~Veerapaneni, J.~D.~Co-Reyes, M.~Chang, et~al.
\newblock Entity Abstraction in Visual Model-Based Reinforcement Learning.
\newblock In \emph{CoRL}, 2020.

\bibitem{wang2024voyager}
G.~Wang, Y.~Xie, Y.~Jiang, A.~Mandlekar, C.~Xiao, Y.~Zhu, L.~Fan, and A.~Anandkumar.
\newblock Voyager: An Open-Ended Embodied Agent with Large Language Models.
\newblock \emph{arXiv:2305.16291}, 2023.

\bibitem{wei2022chain}
J.~Wei, X.~Wang, D.~Schuurmans, et~al.
\newblock Chain-of-Thought Prompting Elicits Reasoning in Large Language Models.
\newblock In \emph{NeurIPS}, 2022.

\bibitem{wertheimer1938laws}
M.~Wertheimer.
\newblock Laws of Organization in Perceptual Forms.
\newblock In W.~D.~Ellis, editor, \emph{A Source Book of Gestalt Psychology}, pages 71--88. Kegan Paul, Trench, Trubner \& Co., 1938.

\bibitem{xu2023arga}
Y.~Xu, S.~Khalil, and S.~Sanner.
\newblock Graphs, Constraints, and Search for the Abstraction and Reasoning Corpus.
\newblock \emph{arXiv:2210.09880}, 2023.

\bibitem{yang2007learning}
Q.~Yang, K.~Wu, and Y.~Jiang.
\newblock Learning Action Models from Plan Examples Using Weighted MAX-SAT.
\newblock \emph{Artificial Intelligence}, 171(2--3):107--143, 2007.

\bibitem{zhuo2019amdn}
H.~H.~Zhuo, J.~Peng, and S.~Kambhampati.
\newblock Action Model Acquisition from Noisy Plan Traces.
\newblock In \emph{IJCAI}, 2019.

\bibitem{grand2020amlsi}
M.~Grand, D.~Pellier, and H.~Fiorino.
\newblock AMLSI: A Novel and Accurate Action Model Learning Algorithm.
\newblock In \emph{KR}, 2020.

\bibitem{aineto2019fama}
D.~Aineto, S.~Jim{\'e}nez~Celorrio, and E.~Onaindia.
\newblock Learning Action Models with Minimal Observability.
\newblock \emph{Artificial Intelligence}, 275:104--137, 2019.

\bibitem{lamanna2021olam}
L.~Lamanna, A.~Saetti, L.~Serafini, A.~E.~Gerevini, and P.~Traverso.
\newblock Online Learning of Action Models for PDDL Planning.
\newblock In \emph{IJCAI}, 2021.

\bibitem{le2024pisam}
H.~S.~Le, B.~Juba, and R.~Stern.
\newblock Learning Safe Action Models with Partial Observability.
\newblock In \emph{AAAI}, 2024.

\bibitem{mordoch2024csam}
A.~Mordoch, E.~Scala, R.~Stern, and B.~Juba.
\newblock Safe Learning of PDDL Domains with Conditional Effects.
\newblock In \emph{AAAI}, 2024.

\bibitem{guan2023llm_pddl}
L.~Guan, K.~Valmeekam, S.~Sreedharan, and S.~Kambhampati.
\newblock Leveraging Pre-trained Large Language Models to Construct and Utilize World Models for Model-based Task Planning.
\newblock In \emph{NeurIPS}, 2023.

\bibitem{wong2024ada}
L.~Wong, J.~Mao, P.~Sharma, Z.~S.~Siegel, J.~Feng, N.~Korneev, J.~B.~Tenenbaum, and J.~Andreas.
\newblock Learning Adaptive Planning Representations with Natural Language Guidance.
\newblock In \emph{ICLR}, 2024.

\bibitem{verma2025instruct}
P.~Verma, N.~La, A.~Favier, S.~Mishra, and J.~A.~Shah.
\newblock Teaching LLMs to Plan: Logical Chain-of-Thought Instruction Tuning for Symbolic Planning.
\newblock \emph{arXiv:2509.13351}, 2025.

\bibitem{garrido2022temporal}
A.~Garrido.
\newblock Learning Temporal Action Models from Multiple Plans: A Constraint Satisfaction Approach.
\newblock \emph{Knowledge-Based Systems}, 252:109410, 2022.

\end{thebibliography}

\end{document}
